---
title: "Predicting Form in Weightlifting - A Coursera Practical Machine Learning Assignment"
author: "akselix"
date: "12 Jun 2015"
output: html_document
---

## Background
This project was part of Coursera course by Joh Hopskins University: [Practical Machine Learning](https://www.coursera.org/course/predmachlearn)

The data for this assignement was collected from on-body sensors and it is available along more information of their study from [Groupware@LES Human Activity Recognition Project](http://groupware.les.inf.puc-rio.br/har#weight_lifting_exercises)

Here is their brief description of the setting:

Six young health participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in five different fashions: exactly according to the specification (Class A), throwing the elbows to the front (Class B), lifting the dumbbell only halfway (Class C), lowering the dumbbell only halfway (Class D) and throwing the hips to the front (Class E). Class A corresponds to the specified execution of the exercise, while the other 4 classes correspond to common mistakes. 

## Assignment
The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases. 



```{r}
summary(cars)
```

You can also embed plots, for example:

```{r, echo=FALSE}
plot(cars)
```

